---
title: "Environmental and Energy Data Analysis Final Assessment"
output:
  word_document: default
  html_document:
    df_print: paged
  html_notebook: default
---


Portfolio submission: Final Analysis Assessment

Course Title: Environmental and Energy Data Analysis

Course Code: 700223_A22_T1

Student ID: 202204752

This report is about statistical analyses done on some sample data collected at 
different times and from different places.

There were three data sets analysed. The processes of analyses and findings  
were all documented as presented in this report.

The three data sets were:

1. The 'penguins' data, 
2. The dorsal_spot_data and
3. The 'crustaceans' data.

#
# SECTION ONE

This section contains the report of the analyses done on the penguins data.

The data was gotten from an R language package called 'palmerpenguins'.

The research question these analyses answered was stated as follows:

Research Question: TO INVESTIGATE IF THE WEIGHT (BODY MASS IN GRAMS) DIFFERS 
                   SIGNIFICANTLY BETWEEN THE THREE PENGUIN SPECIES.
                   

The question was about finding out if the three species of penguin in the data
set differ significantly in terms of their weight(body mass).

The palmerpenguins package was installed and loaded with the codes below.

These gave access to the penguin data used for the analyses.

```{r}
install.packages('palmerpenguins') #to install palmerpenguin package

library(palmerpenguins) #to load palmerpenguin package 
```


After the installing and loading the package, the penguin data was access as 
demonstrated below.


```{r}
peng <- penguins #the penguin data loaded.
```

#

## Data Exploration:

#

```{r}
View(peng) #to view the data loaded.
```

There were 344 observations and 8 variables in the data.

```{r}
names(peng) #to check the variables or columns names in the data.
```

```{r}
str(peng) #to examine the structure of the data.
```

The data contained both numeric and factorial variables.

The body "body_mass_g" was a continuous variable, while the 'species' variable 
was a factor with three levels.


```{r}
summary(peng$body_mass_g) #to check the summary statistics of the body mass.
```

The heaviest penguin had a mass of 6300g, while the lightest had a mass of 2700g.
The mean weight of the penguins was 4202g. The median was 4050g.

25% of them had their weights not more than 3550g. While 75% had their weights
not greater than 4750g.

The data however contained two missing values(NA's) which were later removed 
during data cleaning processes.

```{r}
head(peng, 5) #to the see the first five rows of the data
```

```{r}
tail(peng, 5) #to see the last five rows of the data.
```
#
## Data Wrangling and Cleaning:

To start with, the two columns of interest were filtered out and the NA's removed.

A package called tidyverse was used as part of tools for the analyses.

The package was installed and loaded as shown here:

```{r}
install.packages('tidyverse') #installing 'tidyverse'

library(tidyverse) #loading tidyverse.
```

Tidyverse is a combination of different packages used for statistical analyses.

The codes below were used to filter out the variables of interest and to remove
the missing values.

The pipe operator ( %>% ) was introduced to make the codes and the steps of filtering 
and cleaning shorter. It passed the results of one step to the next.

The select() function was used to filter the variables, the drop_na() function
was used to remove the NA's.

```{r}
dr_peng = peng %>% 
  select(body_mass_g, species) %>% 
  drop_na()
```


```{r}
View(dr_peng) #to view the filtered and cleaned data.
```

```{r}
summary(dr_peng)
```

From the new data, there were 151 individual Adelie species, 68 Chinstraps and 
123 Gentoos.

#
## Analysis of Data

To begin the analysis, the screening tests of the response variable was conducted
because it would determine the type of test to be conducted, whether parametric
or non-parameteric.

1. Checks for Normality:

The code chunk below was used to check the distribution of the body_mass of the 
penguins.
 
```{r}
dr_peng %>%                                 #data set used
  ggplot(aes(sample = body_mass_g))+        #ggplot package from tidyverse.
  stat_qq(colour = 'blue')+                                #for the qqplot
  stat_qq_line()+                           #for the qqline
  labs(title = 'Normal Check with Q-Q plot', #title and label
       y = 'body mass')
  
```

The body mass was non normal when checked visually.

A statistical test was conducted to check the normality.

Null Hypothesis: The body mass of the penguins was normally distributed.

Alternative Hypothesis: The body mass of the penguins was not normally distributed.

```{r}
peng.sh <- shapiro.test(dr_peng$body_mass_g) #Shapiro's test of normality.

peng.sh
```

We rejected the null hypothesis because the p-value was less than 0.05 and concluded
that the data was non normal.

The test statistic = 0.959.


The body mass data was log transformed and normality was checked for once more.

```{r}
log.peng <- log(dr_peng$body_mass_g) #log transformation

```


Null Hypothesis: The logarithm of the body mass data conformed to a normal
                 distribution.
              
Alternative Hypothesis: The logarithm of the body mass data did not conformed to
                        a normal distribution.
                        
```{r}
shapiro.test(log.peng) #shapiro test for normality.
```
The null hypothesis was rejected because the p-value < 0.05. The data was non
normal. The test statistic = 0.97645.


#
# Data Visualisation

The data was visualised with the codes and plot below.

```{r}
dr_peng %>%                           #The data used
  ggplot(aes(species,                 #Independent variable
             body_mass_g,             #Dependent variable
             fill = species))+        #Colour fill specified by the species
  geom_boxplot()+                     #box plot
  labs(title = 'Plot of Body Mass By Species') #Title of the plot.

```
The species appeared to be different in their weights, except for Adelies and 
Chinstraps.

Since the dependent variable failed normality test, and because more than two 
groups were being compared, a non-parametric version of test for comparing more 
than two groups was chosen for the analysis.
                       
#
## KRUSKAL-WALLIS TEST:


The Kruskal-Walli test was used to conduct the test.


Null Hypothesis: There was no significant difference in weight among the three
                 species of penguin.
                 
Null Hypothesis: At least one species differs significantly from the other in
                 terms of their weights.
                 
The test was fitted as described below.
                 
```{r}
kr_peng <- kruskal.test(body_mass_g ~ species, data = dr_peng)

kr_peng
```

RESULTS: The test statistic = 217.6. The degree of freedom was 2. The p-value
         gave a significant value (p-value < 0.05). Therefore, the null hypothesis
         was rejected. 
         
         
CONCLUSION: At least one species of the penguins differed from the others in 
            terms of weight.
            
            
#
## POST HOC TEST:

Kruskal-Wallis's did not show which species differed from the others. Another
test was conducted to see which species differed from the rest.

The Dunn's Test was conducted to determine where the differences were.

The codes below were used to conduct the test and to plot where the differences
were.


```{r}
install.packages('PMCMRplus') #needed for pairwise comparison

install.packages('ggstatsplot') #needed for the dunn Test and the plots.

library(ggstatsplot) #loading the ggstatsplot package.

ggbetweenstats(
  data = dr_peng,                     #data used  
  x = species,                        #explanatory variable
  y = body_mass_g,                    #dependent variable
  type = "nonparametric",             #Kruskal-Wallis
  plot.type = "box",                  #type of plot expected.
  pairwise.comparisons = TRUE,        #to do pairwise comparison
  pairwise.display = "significant"    #to display the significant values.
  
)

```

RESULTS: The p-values for comparing Adelies and Gentoos, and Chinstraps and
         Gentoos were less than 0.05.
         The median of Adelies and Chinstraps were the same(3700g). While the 
         median of Gentoos was 5000g.
         The Overall p-value < 0.05.
         
         
         
CONCLUSIONS: The Adelies did not differ significantly from the Chinstraps. But the 
             Gentoos differed significantly in terms of weight from both the
             Adelies and the Chinstraps.
             
             
             
#*******************************************************************************

#
# SECTION TWO

The report in this section was based on the dorsal spot data. The data was downloaded
from the link below.

https://canvas.hull.ac.uk/courses/64621/files/4017018/download?download_frd=1
#
Research Question: TO INVESTIGATION WHICH CONTINUOUS VARIABLES IN THE DATA HAD 
                   INFLUENCE ON THE SHRIMP SIZE(cl). 
                   
The data was loaded thus after download;

```{r}
dorsal_spot_data <- read.csv(choose.files())
```

It was viewed with code below.

```{r}
View(dorsal_spot_data)
```

Sixteen(16) columns and Ninety-Eight(98) rows were in the data. The only 
continuous variables in the data were:

1. 'rostrum', 
2. 'depth', 
3. 'dsd' - the dorsal spot diameter,
4. 'ed' - the eye diameter, and
5. 'cl' - the carapace length.

These were selected and analysed. The 'cl' being the response variable.

The codes below were used to filtered the columns that were used for the analysis.

```{r}
dorsal <- dorsal_spot_data %>% 
  select(rostrum, depth, dsd, ed, cl)
```


```{r}
View(dorsal) #to view the new data
```

             
There were many missing values in the data. Removing the rows where there were
NA's drastically reduced the sample size. This could make the model not have
enough precision and accuracy.

The NA's were replaced with zero(0) instead as shown below.

```{r}
dorsal[is.na(dorsal)] = 0  #to change the NA's to zeros
```

```{r}
View(dorsal) #Viewed to confirm if NA's had been changed.
```


```{r}
summary(dorsal$cl) #to check the summary statistics of the response variable
```

The carapace length of the shrimps ranged from 0.87mm to 34mm. The mean length
was 6.973 and the median was 4.615. The inter-quartile range was 2.442mm to
9.02mm.

```{r}
plot(dorsal) #plot of the data.
```


SCREENING CHECKS.

Normality assumption according to (amygdala, 2021), (Bartlett, 2013), (Zach, 2022)
(Complete Dessertation, 2013), is only needed the residuals of the model in order
to validate the model. The assumption is not for the response variable.

The response variable 'cl' was log transformed so the residuals could follow a
normal distribution. 

The model fitted with original response variable had its residuals not conforming to
normality.

The dependent variable was transformed as shown below.

```{r}
l_cl <- log(dorsal$cl) #log transformation of cl
```

The log transformed variable was used to fit the model.

Null Hypothesis: The predictor variables did not have a statistically significant
                 relationship with the response variable.
                 
Alternative Hypothesis: The predictor variables had a statistically significant
                        relationship with the response variable.
                        
             
The model was therefore fitted as shown with code chunk below:

```{r}
dorsal_model <- lm(l_cl ~ rostrum + depth + ed + dsd, data = dorsal) #model fitted

summary(dorsal_model) #model summary.
```
RESEULTS: The p-value < 0.05. The null hypothesis was rejected and concluded 
that there was a relationship between the 'cl' and the explanatory variables.

The coefficients of 'rostrum' and 'dsd' were not significant(p-value > 0.05).

The F-statistic = 63.67. The Adjusted R-squared = 0.721. The degree of freedom 
was 93.
#


## MODEL SELECTION.


The Akaike Information Criterion was used for the model selection.

The best model was the model with lowest AIC value. It was gotten by using the 
step() function in both direction as shown.

```{r}
dorsal_step <- step(dorsal_model, direction = c('both')) #stepwise in both direction
```
From the results, the best model was the one with lowest AIC value of 152.24.

Therefore, the best model was the one with the 'rostrum' variable.

```{r}
summary(dorsal_step) #summary of the best model.
```
RESULTS: The intercept as well as the coefficients were significant except for 
         'dsd'.
         The test statistic was 85.79. The Adjusted R-squared = 0.7239.
         The degree of freedom was 94.
         The overall p-value was less than 0.05.
         
CONCLUSIONS: The model could explain 72% of the variations in the log transformed 
             carapace length by the depth, eye diameter and the dorsal spot diameter.



```{r}
dorsal_step #to display the intercept and the coefficients of the best model.
```

Thus, the equation of the model would be: 

l_cl = 0.6185 + 0.811*ed - 0.774*dsd + 0.0006*depth

```{r}

#  l_cl = 0.6185 + 0.811*ed - 0.774*dsd + 0.0006*depth    

```



## MODEL VALIDATION

```{r}
par(mfrow = c(2,2)) #plotting the four plots on a page.

plot(dorsal_step) #the validation plot of the model.

```
           
             
             
CONCLUSION: Since the Normal Q-Q plot showed normality, thus the model was valid.

```{r}
shapiro.test(dorsal_step$residuals) #Shapiro's test of the residual for normality
```

The p-value > 0.05. Thus the model was valid.  


#
#*******************************************************************************
#

# SECTION THREE

This third section of report is based on the analyses conducted on the 'crustaceans'
data. The data was downloaded from the link below.

https://canvas.hull.ac.uk/courses/64621/files/4225671/download?download_frd=1

The data was about some crustaceans sampled over a period of three(3) years from
four different sites.

Research Question 1: ARE THERE SIGNIFICANT DIFFERENCES IN THE PROPORTIONS OF MALE 
                     AND FELAME CRUSTACEANS IN THE YEARS?
                   

Research Question 2: IS THE SIZE SIGNIFICANTLY DIFFERENT AMONG THE SPECIES OF THE 
                     CRUSTACEANS?
                     

The data was downloaded and loaded into R as described here.

```{r}
agaza = read.csv(choose.files()) #loading the data file.
```


The data was viewed with code chunk below.

```{r}
View(agaza) #viewed the data.
```

The sample data was quite large. It contained 61,737 observations and 8 variables.


```{r}
str(agaza) #the structure of the data.
```

It consisted of only one continuous variable 'Size', one discrete variable 'Year'
and six(6) categorical variables.

To answer the first question, the 'Year' variable was converted to a factor and a
contingency or frequency table built as shown.

```{r}
fYear <- factor(agaza$Year) #converting 'Year' to a factor .
```


```{r}
agaza_t <- table(agaza$Sex, fYear) #contingency table

View(agaza_t) #viewed the table
```

The Chi-squared test of independence was used to evaluate the variations in sex
of the crustaceans from year to year.


```{r}
#A plot showing the variations

ggplot(data.frame(agaza_t),               #table converted to dataframe
       aes(fYear, Freq, fill = Var1))+    #Aesthetics from the frequency table.
  geom_bar(stat = 'identity')+            #Type of plot
  facet_grid(rows = vars(Var1))           #To separate by sex
```



Null Hypothesis: There was no significant difference in proportions of male and
                 female crustaceans from year to year.
                 
Alternative Hypothesis: There was a significant difference in proportions of male 
                        and female crustaceans from year to year.


The Chi-squared test was fitted as follows:

```{r}
chisq <- chisq.test(agaza_t) #chi-squared test

chisq
```

RESULTS: The test statistic = 436.91. The degree of freedom was 2. The p-value
         was less than 0.05.
         
CONCLUSIONS: The null hypothesis was rejected and it was concluded that there was
             a significant difference in the populations of male and female
             crustaceans from 2013 to 2017.
             
             
```{r}
chisq$observed #observed counts
```


```{r}
round(chisq$expected, 2) #expected counts to 2 significant figures.
```

```{r}
round(chisq$residuals, 2) #residuals of the test to 2 significant figures.
```

The residuals were visualised using the 'corrplot' package as described here. 

```{r}
install.packages('corrplot') #installing the 'corrplot' package.

library(corrplot) #importing the 'corrplot' package.

corrplot(chisq$residuals, is.cor = FALSE) #plotting residuals of the test.

```


RESULTS: 

The plot above gave a summary of the variations between expected and the observed 
populations of crustaceans.
         
The red balls represented a fall short from the expected values, and their sizes
represented the magnitude of the fall.

The blue balls represented an increase over the expected values. Their sizes
depicted the magnitude of the increase.


INTERPRETATIONS AND CONCLUSIONS:

At the start of the sampling years, the female crustaceans were less than what 
was expected. But over the years, the observed populations were more than expected.

The male crustaceans started with observed population being more than the expected.
But as time went on, the population of male crustaceans went below the expected.
This could be as resulted of migration, death, fishing activities or migration.

Overall, the total population of crustaceans reduced as year passed by.

Conservation efforts need to be taken to prevent the crustaceans from going into
extinction.
             
             
Answer to Research Question 2:

To answer this question, the Size and the Species were examined. They were 
summarised as done below.



```{r}
#These codes grouped the crustaceans by species with their mean sizes.
agaza %>% 
  group_by(Species) %>% 
  summarise('Average Size' = mean(Size))

```

The analysis was to find out if difference in the means are statistically 
significant or not.

Because we had a large sample size, normality was assumed (Complete Dessertation, 2013).

Homogeneity of variance test was conducted as follow.


Null Hypothesis: The variances of the groups are the same.

Alternative Hypothesis: The variances of the groups are not the same.

```{r}
bartlett.test(Size ~ Species, data = agaza) #Bartlett test of homogeneity.
```
The test statistic was 17677, The degree of freedom was 2. The p-value < 0.05.

Thus, we rejected the null hypothesis and concluded that the variances are not the
same.

A parametric test ANOVA was conducted to find out if the differences in the means 
were significant.


```{r}
table(agaza$Species) #table to check distribution of species.
```

The distribution showed we had an unbalanced design data. Therefore, a type 'III'
anova was conducted. 

Accoording to R documentation for aov(), it is was built only for balanced designs.

Null Hypothesis: There was no statistically significant difference in the means
                 sizes of the sexes of the crustaceans.
                 
Alternative Hypothesis: At least the mean size of one species of the crustaceans
                        statistically differed from the others.
                        
                        

```{r}
agaza_a <- aov(Size ~ Species, data = agaza) #ANOVA fitted.


summary(agaza_a) #summary of the ANOVA.
```

The degree of freedom for the Species was 2, and for the residuals was 61734.

The test statistic was 86459. The p-value < 0.05. Thus, the null hypothesis was
rejected. It was concluded that the species differed significantly in size.


```{r}
install.packages('car') #installing the car package.


library(car) #load the car package.


Anova(agaza_a, type = 'III') #Type III anova fitted.
```

The p-value < 0.05. Thus it was concluded that there statistically significant
difference in the sizes of the species.


The codes below was used to visualised the data set.

```{r}

my_fill = c('#4e84c4', '#e69f00', '#c4849f')  #To set custom fill
#                                              for the bar representing each group

ggplot(agaza,                                                 #data used in the plot
       aes(Species, Size, fill = Species))+                   #variables used for the plot
  scale_fill_manual(values = my_fill)+                        #applying the custom fill
  geom_bar(stat = 'summary',                                  #summary statistics
           fun = 'mean',                                      #the means
           width = 0.6)+                                      #the width of bars
  theme_bw()+                                                 #black-white theme
  labs(title = 'Plot of Size Against Species of Crustaceans') #the title of the plot

```

From the plot above, the Crabs were the bigger species of the crustaceans, followed
by the Lobsters and the Velvets. The Velvets were the smallest in size.
#

## Post Hoc Test

TukeyHSD test was used to conduct a post hoc test to see where the differences are.


```{r}

tuk_agaza = TukeyHSD(agaza_a) #TukeyHSD test.

tuk_agaza #Tukey summary.

```


```{r}
plot(tuk_agaza) #Tukey plot for validation

```


From the post hoc test results, all the p adj values were zeros. This implied
That all the species were different from one another in terms of sizes.



#
# LIST OF REFERENCES

Patil, I. (2021). Visualizations with statistical details: The 'ggstatsplot' approach.
Journal of Open Source Software, 6(61), 3167, doi:10.21105/joss.03167


an amygdala, 2021. What Is the Assumption of Linearity in Linear Regression?
https://medium.com/the-data-base/what-is-the-assumption-of-normality-in-linear-regression-be9f06dae360


Jonathan Bartlett, 2013. Assumptions for linear regression: https://thestatsgeek.com/2013/08/07/assumptions-for-linear-regression/


Zach, 2020. The Four Assumptions of Linear Regression: https://www.statology.org/linear-regression-assumptions/


Statistics Solutions. (2013). Normality . Retrieved from https://www.statisticssolutions.com/academic-solutions/resources/directory-of-statistical-analyses/normality/
         
R Core Team (2022). R: A language and environment for statistical computing. 
R Foundation for Statistical Computing, Vienna, Austria. 
URL https://www.R-project.org/.


Wickham H, Averick M, Bryan J, Chang W, McGowan LD, François R, Grolemund G, 
Hayes A, Henry L, Hester J, Kuhn M, Pedersen TL, Miller E, Bache SM, Müller K,
Ooms J, Robinson D, Seidel DP, Spinu V, Takahashi K, Vaughan D, Wilke C, Woo K,
Yutani H (2019). “Welcome to the tidyverse.” _Journal of Open Source Software_,
*4*(43), 1686. doi:10.21105/joss.01686 <https://doi.org/10.21105/joss.01686>.


John Fox and Sanford Weisberg (2019). An {R} Companion to Applied Regression, 
Third Edition. Thousand Oaks CA: Sage. 
URL: https://socialsciences.mcmaster.ca/jfox/Books/Companion/


Ogle, D.H., J.C. Doll, P. Wheeler, and A. Dinno. 2022. FSA: Fisheries Stock Analysis. 
R package version 0.9.3, https://github.com/fishR-Core-Team/FSA.
         
         